Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-29 manuscript/manuscript.tex
Minor edit on Algorithm performance
Review the new performance section

Because the density-based discretization algorithm introduces more division along the
radial dimension, it is reasonable that the computation time for variable densities will
be increased in comparison to the homogeneous density case.
In order to get a qualitative indicator of this increase in the computation time we
gathered the number of subdivisions the density-based discretization performs under
exponential and sinusoidal density cases.

We assign an exponential density (Eq.~\ref{eq:density-exp}) with values of $b$ equal to
1, 2, 5, 10, 30 and 100, and a sinusoidal density (Eq.~\ref{eq:density-sine}) with
values of $b$ equal to 1, 2, 5, and 10 to an arbitrary tesseroid.
Then we applied the density-based discretization algorithm on each case recording how
many subdivisions it carried out and where the boundaries of the smaller tesseroids are
located.
Figs.~\ref{fig:number-of-tesseroids}(a) and~\ref{fig:number-of-tesseroids}(b) show the
exponential and sinusoidal densities used and the resulting boundaries of the
discretized tesseroids as orange dots.
On both exponential and sinusoidal density cases a value of $\delta$ equal to 0.1 has
been used.

On the exponential density scenario, the algorithm performs a single split independently
of the value of $b$, i.e.~the number of discretizations is independent of the
variability of the exponential function.
Therefore, the computation time would be increased in comparison to the homogeneous
density case due to this single split, but would be independent of the value of $b$.

On the other hand, the sinusoidal density case shows a strong dependence of the
number of discretizations on the value of $b$.
Fig.~\ref{fig:number-of-tesseroids} shows an almost linear relation between these two
magnitudes.
This means that the computation time is heavily bounded to the value of $b$ used.
It can be greatly increased for large values of $b$ such as 100.

From this results we can conclude that the computation time would be greatly increased
in case of multiple inflection points on the density function.
Nevertheless, for most geophysical applications on which the density profiles may have
low number of inflection points, the computation time would be in the same order of
magnitude as the one for the homogeneous density case.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/number-of-tesseroids.pdf}
\caption{
    Number of discretizations performed by the density-based algorithm in case of
    (a)~exponential densities with the same values of $b$ shown in
    Fig.~\ref{fig:exp-densities},
    (b)~sinusoidal densities with the same values of $b$ shown in
    Fig.~\ref{fig:sine-densities}.
    On both figures the boundaries of the discretized tesseroids are marked with orange
    dots.
    Subfigure (c) shows the number of discretized tesseroids for the sinusoidal
    densities as a function of the factor $b$.
}
\label{fig:number-of-tesseroids}
\end{figure*}

@@

>>>>

Because the density-based discretization algorithm introduces more divisions along the
radial dimension, it is reasonable that the computation time for variable densities will
be higher than the homogeneous density case.
Directly comparing the computation time of both cases cannot be done in a meaningful
way because it is highly dependent on the implementation of the algorithm,
the choice of programming language, and the particular density function used.
In order to obtain an indicator of the increase in the computation time,
we chose the number of tesseroid divisions in the density-based discretization as a
proxy measure.

We analyse exponential density functions (Eq.~\ref{eq:density-exp}) with values of $b$ equal to
1, 2, 5, 10, 30 and 100, and sinusoidal density functions (Eq.~\ref{eq:density-sine}) with
values of $b$ equal to 1, 2, 5, and 10.
We then apply the density-based discretization algorithm on each function and record the
number divisions carried out.
Figs.~\ref{fig:number-of-tesseroids}(a) and~\ref{fig:number-of-tesseroids}(b) show the
exponential and sinusoidal density functions and the resulting discretization points as
orange dots.

For the exponential density functions, the algorithm performs a single division on the
inflection point independently of the value of $b$.
Therefore, the computation time would likely be at least twice that of the homogeneous
density case (assuming equal implementations).
On the other hand, Fig.~\ref{fig:number-of-tesseroids}c shows an almost linear relation
between the number of divisions in the sinusoidal density case and the value of $b$.
Thus, the computation time is likely to be dependent on the number of wavelengths
of the sinusoidal function that are contained within the tesseroid.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/number-of-tesseroids.pdf}
\caption{
    Number of divisions performed by the density-based discretization algorithm
    (with $\delta = 0.1$) in case of
    (a)~exponential density functions with the same values of $b$ shown in
    Fig.~\ref{fig:exp-densities},
    (b)~sinusoidal density functions with the same values of $b$ shown in
    Fig.~\ref{fig:sine-densities}.
    On both figures, the locations of the tesseroid divisions are are marked with orange
    dots.
    (c) shows the number of discretized tesseroids for the sinusoidal density functions
    as a function of $b$.
}
\label{fig:number-of-tesseroids}
\end{figure*}

@@

The algorithm performance analysis shows that the computation time using variable
densities is likely to be at least two times larger than the homogeneous density case.
It is also likely to increase proportionally to the number of inflection points of the
density function.
As is the case for most numerical methods, there is a trade-off between computation time
and accuracy.
Nevertheless, density profiles will have few inflection points for most geophysical
applications.
Therefore, the computation time would be in the same order of
magnitude as the one for the homogeneous density in most real world applications.
